\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping

### 2018-08-04 revision/response.tex

\EditorQuestion{
   For instance, one of the other papers could provide the benchmark data that is required for [the software] manuscript.
   So the software paper can include those files and refer to the other manuscript for details.
}
\Response{
  We have incorporated this recommendation, using the dataset from Kumar et al. 2013 that was used in one of the case studies in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
  The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``Performance Benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
}

\subsection*{Issues Specific to this Manuscript}

\EditorQuestion{
  As suggested by Reviewer 1 it will be very useful to have a benchmark study with times/memory requirements as a function of number of markers and sample size.
  This study will give those interested on using the package an understanding of how this package would scale.
  It will be important to include in the discussion also elements related to scalability: is the package suitable for very large-p/large-n? 
}
\Response{
  We have added this information in the ``Performance Benchmarks'' section. 
  See detailed response to Reviewer 1 comment 3. 
  Note that this has lengthened the paper somewhat, but we can move plots to supplemental if that would be preferred.
}

\EditorQuestion{
  There is a very rich literature in the study of genetic control in environmental variance in both plant and animal breeding as well as model organisms.
  This literature should be mentioned. This comment may also apply to some of the companion papers.
  [Citation recommendations provided by the editor.]
}
\Response{
  Thank you for these suggestions. 
  This literature was foundational toward the development of the work presented in this manuscript, and we agree with the editor that it should be cited in the introduction.
  We have improved the introduction to describe how recognizing the heritability of environmental variance is a necessary step toward vQTL and mvQTL mapping and used many of the suggested citations in this section.
}

@@

% Ibáñez N, Varona L, Sorensen D, Noguera JL. A study of heterogeneity of environmental variance for slaughter weight in pigs. Animal. 2007;2:19-26.

% Sørensen P, de los Campos G, Morgante F, Mackay TFC, Sorensen D. Genetic Control of Environmental Variation of Two Quantitative Traits of Drosophila melanogaster Revealed by Whole-Genome Sequencing. Genetics. 2015;201(2):487-497. doi:10.1534/genetics.115.180273.

\section*{Reviewer 1 Comments}

\ReviewerQuestion{
  The authors mention ``in almost all traits of interest in human health and disease''.
  The detection of QTL has also been widely used in plant and animal populations.
}
\Response{
  The reviewer is quite correct that QTL mapping has been productively applied in plant an animal populations in addition to human populations.
  Rather than add references from plant and animal genetics to the introduction, we recognized that the introduction started unnecessarily broadly and therefore removed the first sentence.
  To address the reviewer's concern, however, we have added references to the next paragraph in the introduction (where the heritability of phenotype variance is discussed) that rely heavily on studies of model organisms, livestock and crops.
}

\ReviewerQuestion{
  The authors used a set of simulated data with 3 chromosomes and 11 equally spaced markers.
  I would like to see an example with a larger genotype data set.
}
\Response{
  The unrealistically-small genome we used to illustrate the standard workflow serves an important purpose --- to allow a reader to replicate the procedure and its ``findings'' on a typical desktop in a matter of minutes.
  Since this manuscript has been submitted as part of a three-paper companionship where each of the other papers contains analyses of real data using package \texttt{vqtl}, we feel that using a reduced dataset to demonstrate the package functions is reasonable in this instance.
  Additionally, in response to other reviewer comments, we have added performance benchmarks to this manuscript, which make use of the dataset from Kumar et al 2013, as described below.
}

### 2018-08-04 Package_vqt_Manuscript.tex
finished round 1 of revisions
Signed-off-by: Robert Corty <rcorty@gmail.com>

Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
%
%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\CortyMethodsPaper),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
%
%The R package \texttt{vqtl} implements mean-variance QTL mapping
which is currently suitable for F2 intercrosses and backcrosses, and is inter-operable with the well-established mean QTL-oriented package for this purpose, \texttt{qtl} \citep{Broman2003}.
%We introduce the \texttt{vqtl} package as follows.
First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
The association statistics of all four tests are initially plotted in LOD score units, with drawbacks of this plotting unit discussed; then permutation scans are used to determine empirically-adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more easily comparable.
Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
Last, we benchmark performance, using one of the datasets examined in \CortyReanalysisPaper to report how computation time varies with marker density and number of permutations.

@@

To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using \texttt{R/qtl} package \citep{Broman2003}, on which \texttt{vqtl} is based.
This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden Markov model. We then generated four phenotypes:
\begin{enumerate}
	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests.
	\item \texttt{phenotype2} has an mQTL that explains 4\% of phenotype variance at the center of chromosome one.
	\item \texttt{phenotype3} has a vQTL at the center of chromosome two.
        This vQTL acts additively on the log standard deviation scale, and results in residual standard deviation of [0.8, 1, 1.25] for the three genotype groups.
	\item \texttt{phenotype4} has an mvQTL at the center of chromosome three.
        This mvQTL has a mean effect that explains 2.7\% of phenotype variance and a variance effect that acts additively on the standard deviation scale, resulting in residual standard deviation of [0.85, 1, 1.17] for the three genotype groups.
\end{enumerate}

@@

The central function for genetic mapping in package \texttt{R/qtl} is \texttt{scanone} \citep{Broman2003}.
Analogously, the central function for mean-variance QTL mapping in package \texttt{vqtl} is \texttt{scanonevar}, building on an early version of \texttt{scanonevar} in package \texttt{qtl}.
It takes three required inputs:
\begin{enumerate}
    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an experimental cross, as defined in package \texttt{qtl}.
    \item \texttt{mean.formula} is a two-sided formula, specifying the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted, with keywords \texttt{mean.QTL.add} and \texttt{mean.QTL.dom}
    \item \texttt{var.formula} is a one-sided formula, specifying the variance covariates to be corrected for as well as the QTL terms to be fitted, using keywords \texttt{var.QTL.add} and \texttt{var.QTL.dom}.
\end{enumerate}
% Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
% Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
% This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
% \RC{}{Doesn't fit.  It's already in the online code. Delete?}
For example, to scan a phenotype named \texttt{p1}, we run:
\begin{verbatim}
scanonevar(
 cross = test_cross,
 mean.formula = p1 ~ sex + mean.QTL.add + mean.QTL.dom,
 var.formula = ~ sex + var.QTL.add + var.QTL.dom
)
\end{verbatim}
At each locus in turn, this function tests for the presence of an mQTL, a vQTL, and an mvQTL. The basis of these tests is a comparison between the fit of an alternative model of the form
\begin{align*}
    \text{mean} &= \text{covariate effects} + \text{locus effects}\\
    \log(\text{variance}) &= \text{covariate effects} + \text{locus effects}    
\end{align*}
with a null model that omits specific terms: for the mQTL test, the null model omits locus effects on phenotype mean; for the vQTL test, the null omits the locus effects on phenotype variance; and for the mvQTL test, the null omits locus effects on both mean and variance.
(Note that the mQTL test in mean-variance QTL mapping is different from the traditional test: the traditional test does not have variance predictors of any kind in either null or alternative models.)

@@

Based on these benchmarks, the workflow presented here is practical for QTL mapping F2 intercross and similar populations on modern, multi-core scientific computers.
Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci, could not be practically analyzed with package \texttt{vqtl} in this way, although it is likely that statistical and computational steps could be taken to make such studies feasible:
statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004};
computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.

>>>>

Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
% \WV{}{Should we be using ``R/vqtl'' and ``R/qtl'' terminology?}
%
%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\CortyMethodsPaper),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
%
%The R package \texttt{vqtl} implements mean-variance QTL mapping
which is currently suitable for F2 intercrosses and backcrosses, and is inter-operable with the well-established mean QTL-oriented package for this purpose, \texttt{qtl} \citep{Broman2003}.
%We introduce the \texttt{vqtl} package as follows.
First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
The association statistics of all four tests are initially plotted in LOD score units, with drawbacks of this plotting unit discussed; then permutation scans are used to determine empirically-adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more easily comparable.
Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
Last, we benchmark performance, using one of the datasets examined in \CortyReanalysisPaper to report how computation time varies with marker density and number of permutations.

@@

To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using the \texttt{R} package {qtl} \citep{Broman2003}, on which \texttt{vqtl} is based.
This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{qtl}'s hidden Markov model. We then generated four phenotypes:
\begin{enumerate}
	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests.
	\item \texttt{phenotype2} has an mQTL that explains 4\% of phenotype variance at the center of chromosome one.
	\item \texttt{phenotype3} has a vQTL at the center of chromosome two.
        This vQTL acts additively on the log standard deviation scale, and results in residual standard deviation of [0.8, 1, 1.25] for the three genotype groups.
	\item \texttt{phenotype4} has an mvQTL at the center of chromosome three.
        This mvQTL has a mean effect that explains 2.7\% of phenotype variance and a variance effect that acts additively on the standard deviation scale, resulting in residual standard deviation of [0.85, 1, 1.17] for the three genotype groups.
\end{enumerate}

@@

The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone} \citep{Broman2003}.
Analogously, the central function for mean-variance QTL mapping in package \texttt{vqtl} is \texttt{scanonevar}, building on an early version of \texttt{scanonevar} in package \texttt{qtl}.
It takes three required inputs:
\begin{enumerate}
    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an experimental cross, as defined in package \texttt{qtl}.
    \item \texttt{mean.formula} is a two-sided formula, specifying the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted, with keywords \texttt{mean.QTL.add} and \texttt{mean.QTL.dom}
    \item \texttt{var.formula} is a one-sided formula, specifying the variance covariates to be corrected for as well as the QTL terms to be fitted, using keywords \texttt{var.QTL.add} and \texttt{var.QTL.dom}.
\end{enumerate}
% Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
% Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
% This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
% \RC{}{Doesn't fit.  It's already in the online code. Delete?}
For example, to scan a phenotype named \texttt{p1}, we run:
\begin{verbatim}
scanonevar(
 cross = test_cross,
 mean.formula = p1 ~ sex + mean.QTL.add + mean.QTL.dom,
 var.formula = ~ sex + var.QTL.add + var.QTL.dom
)
\end{verbatim}
At each locus in turn, this function tests for the presence of an mQTL, a vQTL, and an mvQTL. The basis of these tests is a comparison between the fit of an alternative model of the form
\begin{align*}
    \text{mean} &= \text{covariate effects} + \text{locus effects}\\
    \log(\text{variance}) &= \text{covariate effects} + \text{locus effects}    
\end{align*}
with a null model that omits specific terms: for the mQTL test, the null model omits locus effects on phenotype mean; for the vQTL test, the null omits the locus effects on phenotype variance; and for the mvQTL test, the null omits locus effects on both mean and variance.
(Note that the mQTL test in mean-variance QTL mapping is different from the traditional test: the traditional test does not have variance predictors of any kind in either null or alternative models.)

@@

Based on these benchmarks, the workflow presented here is practical for QTL mapping F2 intercross and similar populations on modern, multi-core scientific computers.
Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci, could not be practically analyzed with package \texttt{vqtl} in this way, although it is likely that statistical and computational steps could be taken to make such studies feasible:
statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004};
computationally, the software could be modified to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
